SoK: Application Attack Defenses (in Question)

### 2014-09-23 docs/questioning.tex
better definitions and discussion

The evaluation of

\section{Analysis Dimensions}

\begin{itemize}
\item {\bf Sensitivity}, the probability that an attack will be
  caught.
\item {\bf Specificity}, the probability that a normal interaction
  will test negative.
\item {\bf Positive Predictive Value} (PPV), the probability that a
  reported attack is a real attack.
\item {\bf Negative Predictive Value} (NPV), the probability that if
  nothing is reported no attack has taken place.
\end{itemize}

To do any meaningful calculation of the above, we need to define
precisely the following in our methodology:
\begin{itemize}
\item True Positive (TP)
\item False Positive (FP)
\item True Negative (TN)
\item False Negative (FN)
\end{itemize}

\section{Application Attack Defense Mechanisms Accuracy}
% @Panos: we need to change our terminology. Instead of intrusion-detection
% systems we need to move to something like application attack defences.
% Intrusion detection is very broad and we are currently
% surveying mechanisms that detect attacks coming from a specific catefory.

An intrusion detection system examines some software artifact and
reports vulnerabilities. The value an intrusion detection system can
be measured in terms of its \emph{accuracy}, which gauges the quality
of its output. An intrusion detection system:
\begin{itemize}
\item A prediction about a vulnerability is true. That is a \emph{true
    positive}. 
\item A prediction about a vulnerability is false; intrusion
  detection reported something that in reality is not vulnerability.
  That is a \emph{false positive}.
\item A prediction about the absence of a vulnerability is true. That
  is a \emph{true negative}: intrusion detection evaluated something
  as being safe, and it is.
\item A prediction about the absence of a vulnerability is false. That
  is a \emph{false negative}: intrusion detection missed a
  vulnerability.
\end{itemize}
In terms of the above, the \emph{Positive Predictive Value} (PPV) is
defined as the ratio of true positives over all reported positives,
while the \emph{Negative Predictive Value} (NPV) is defined as the
ratio of true negatives over all reported negatives. In intrusion
detection, the PPV is an indicator of how much we should be worried
about a reported vulnerability being true. Conversely, the NPV is an
indicator of how much we should be reassured when no vulnerabilities
are reported.

What we need to proceed:
\begin{itemize}
\item Do the studies report the sample size? Note that we must be
  careful with what we mean by sample size. A single item in the
  sample is something that can be vulnerable or not. 
\item Do the studies report true positives, false positives, true
  negatives, false negatives?
\end{itemize}

>>>>

\section{Methodological Background}

As in medical diagnostic tests, application attack defense mechanisms
must demonstrate the presence of an attack. This, however, does not
make an application attack defense mechanism immediately useful. In
order for a detection mechanism to be useful it must accurate, easy to
use, and economical. The accuracy of a detection mechanism is gauged
with the following metrics:
\begin{itemize}
\item {\bf Sensitivity}, the probability that an attack will be
  caught.
\item {\bf Specificity}, the probability that a normal interaction
  will test negative.
\item {\bf Positive Predictive Value} (PPV), the probability that a
  reported attack is a real attack.
\item {\bf Negative Predictive Value} (NPV), the probability that if
  nothing is reported no attack has taken place.
\end{itemize}

For example, a simple announcement that an attach defense mechanism
catches an attack says very little about the usability and
applicability of the mechanism. There may be some value in indicating
that detection \emph{per se} is possible, if that has been considered
infeasible in the past. Most value, though, comes from knowing the
answers to questions like the following, which correspond to the
previous metrics:
\begin{itemize}
\item Does it catch all the attacks? If not, how many of them?
\item Does it catch only attacks? If not, how many false alarms would
  we expect?
\item How much should we be worried if we do get an alarm?
\item How much at ease should we be if no alarm is raised?
\end{itemize}

Sensitivity, specificity, PPV, and NPV, are defined precisely by way
of the $2\times 2$ Table~\ref{tab:attacks-outcomes}. In the cells of
the table we distinguish:
\begin{itemize}
\item True Positive (TP), a real attack that raises an alarm.
\item False Positive (FP), an event that although it is not an attack
  raises an alarm.
\item True Negative (TN), an event that is not an attack and that does
  not raise an alarm.
\item False Negative (FN), an event that although is as attack does
  not raise an alarm.
\end{itemize}

With these we can calculate:

\begin{equation}
\textrm{sensitivity} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{TN}}
\end{equation}

\begin{equation}
\textrm{specificity} = \frac{\textrm{TP}}{\textrm{TN} + \textrm{FP}}
\end{equation}

\begin{equation}
\textrm{PPV} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}}
\end{equation}

\begin{equation}
\textrm{NPV} = \frac{\textrm{TN}}{\textrm{TN} + \textrm{FN}}
\end{equation}

\begin{table}[ht]

\caption{Attacks and Test Outcomes}
\label{tab:attacks-outcomes}

\begin{tabular}{l|l|c|c|}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No \\ \cline{2-4}
\multirow{2}{*}{Result} &  Yes &  True Positive & False Positive \\
& No & False Negative & True Negative \\ \cline{2-4}
\end{tabular}

Sensitivity and specificity can be calculated based on test results
alone. To calculate the sensitivity, we run the test on a controlled
environment where we allow only attack events to reach the system. The
ratio of reported attacks over all events will give us the
sensitivity. Similarly, to calculate the specificity we can run the
test on a controlled environment where we allow only innocuous events
to reach the system. The ratio of non-reported events over all events
will give us the specificity.

The calculation of PPV and NPV requires that we evaluate the attack
defense mechanism on a real operational environment where the events
are a mix of attacks and innocuous events and, moreover, the mix ratio
is the ratio of attacks over harmless events that exists in actual
operations. Too few or too many attacks will skew metrics calculations.

The last requirement may seem cumbersome, but consider that, in actual
operations, PPV is what would determine our actions. Administrators
will limited resources need to allocate their attention judiciously.
They may need to prioritize responses to emergency warnings; the PPV
value will indicate them what priority to give to a raised alarm.
